
---
title: "Multi-City Fixed Effects DiD with Placebo & Parallel Trends Diagnostics"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, broom, lmtest, fixest, sandwich, here, modelsummary)

```

## Overview

This analysis evaluates the effect of decriminalization in Eugene, Oregon using a Difference-in-Differences (DiD) design extended to multiple cities and outcomes. It includes fixed effects, placebo tests, and parallel trends diagnostics.

## Simulated Panel Data

```{r data}

eug <- read.csv(here("data/Provisional Mortality Statistics, 2018 through Last Week Eugene.csv"))
bois <- read.csv(here("data/Provisional Mortality Statistics Boise, 2018 through Last Week.csv"))
```

```{r}
# Keep only Year and Crude Rate, and add a County label
eug_clean <- eug %>%
  select(Year, Crude.Rate) %>%
  mutate(County = "Lane") %>% 
  drop_na()

bois_clean <- bois %>%
  select(Year, Crude.Rate) %>%
  mutate(County = "Ada") %>% 
  drop_na()

combined_data <- bind_rows(eug_clean, bois_clean) %>%
  rename(DeathRate = Crude.Rate) %>%
  mutate(
    Post = ifelse(Year >= 2021, 1, 0),
    Treat = ifelse(County == "Lane", 1, 0),
    DID = Post * Treat
  )
```

## Parallel Trends Check

```{r trends}
# set.seed(123)
# data <- expand.grid(
#   city = c("Eugene", "Pocatello", "Nampa", "Boise"),
#   year = 2017:2022
# ) %>%
#   mutate(
#     treatment_group = ifelse(city == "Eugene", 1, 0),
#     post_policy = ifelse(year >= 2021, 1, 0),
#     did = treatment_group * post_policy,
#     overdose_rate = 30 + 3 * treatment_group - 2 * post_policy - 5 * did + rnorm(n(), 0, 3),
#     arrest_rate = 140 - 4 * treatment_group + 2 * post_policy + 5 * did + rnorm(n(), 0, 5),
#     treatment_admissions = 190 + 10 * treatment_group + 5 * post_policy + 15 * did + rnorm(n(), 0, 10)
#   )
### from simulation, idea for other analysis 

# data_long <- data %>%
#   pivot_longer(cols = c(overdose_rate, arrest_rate, treatment_admissions),
#                names_to = "outcome", values_to = "value")
# 
# ggplot(data_long, aes(x = year, y = value, color = city)) +
#   geom_line() +
#   facet_wrap(~ outcome, scales = "free_y") +
#   labs(title = "Parallel Trends Check (Pre-Treatment Years)", y = "Value", x = "Year") +
#   geom_vline(xintercept = 2021, linetype = "dashed")+
#   theme_minimal()

combined_clean <- combined_data %>%
  filter(!grepl("provisional", Year)) %>%
  mutate(Year = as.numeric(as.character(Year)))

# Filter pre-treatment years (before 2021)
pre_data <- combined_clean %>% filter(Year < 2021)

# Plot pre-trends only
paral_trends <- ggplot(pre_data, aes(x = Year, y = DeathRate, color = County)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Pre-Policy Trends in Overdose Death Rates",
    subtitle = "Lane vs. Ada County (2018–2020)",
    x = "Year",
    y = "Deaths per 100,000"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
print(paral_trends)
ggsave(here("figures_tables/paral_trends.png"))
## Parallel trends assumption 


```


```{r}


# ==== Visualize the Trends with Intervention Line ====
trends_plot <- ggplot(combined_data, aes(x = as.numeric(Year), y = DeathRate, color = County)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = 2021, linetype = "dashed", color = "gray30") +
  annotate("text", x = 2021.2, y = max(combined_data$DeathRate, na.rm = TRUE) - 1,
           label = "Measure 110", hjust = 0, size = 4) +
  labs(
    title = "Crude Overdose Death Rates (2018–2023)",
    subtitle = "Lane County (Eugene) vs. Ada County (Boise)",
    x = "Year",
    y = "Deaths per 100,000 Population",
    color = "County"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    legend.position = "bottom"
  )

print(trends_plot)

ggsave(here("figures_tables/trends_plot.png"))
```



```{r}
# summary statistics
# Convert Year to numeric and filter out provisional 2024 rows

# Pivot to wide format: one column for each county's death rate
wide_data <- combined_clean %>%
  select(Year, County, DeathRate) %>%
  pivot_wider(names_from = County, values_from = DeathRate)

# View the reshaped dataset
print(wide_data)
library(modelsummary)

datasummary_skim(wide_data,
                 type = "all",
                 fmt = 2,
                 fun_numeric = list(Mean = Mean, SD = SD, Min = Min,  Max = Max),
                 output = here("figures_tables/sum_stats.docx"))

# 
# skimr::skim(wide_data)


## citation("fixest")
```

## Fixed Effects DiD Models

```{r models}

# ==== OLS DiD Regression ====
did_model <- lm(DeathRate ~ Post + Treat + DID, data = combined_clean)
summary(did_model)

# What this model does:
# 	•	Post: Captures change over time across all units
# 	•	Treat: Captures baseline difference between Lane and Ada
# 	•	DID: Captures the DiD estimate — the treatment effect
# 
#  Interpretation:
# 	•	DID coefficient = estimated causal effect of Measure 110 on overdose rates
# 	•	Check p-value for significance
# 	•	Examine R² to see model fit

# ==== Fixed Effects DiD Regression ====
did_fe <- feols(DeathRate ~ Treat * Post | Year, data = combined_clean)
summary(did_fe)

#	•	Controls for time, but does not absorb differences between counties.
# 	•	Treat * Post expands to: Treat, Post, DID
# 	•	| County + Year adds fixed effects:
# 	•	County FE: controls for all time-invariant differences (e.g., baseline mortality rates)
# 	•	Year FE: controls for common shocks (e.g., national trends in fentanyl use)
# 
# Interpretation:
# 	•	The coefficient on Treat:Post (aka DID) is my treatment effect
# 	•	More robust to omitted variable bias than basic OLS
# 	•	Standard errors are better estimated with clustered SEs 

modelsummary(list(
  `OLS Regression Model` = did_model,
  `Fixed Effects Model` = did_fe),
             fmt = 2,
               statistic = "conf.int",
             stars = TRUE) 
  # output = here("figures_tables/mod_output.docx"))

```

```{r}
# Add Clustered Standard Errors
## The standard error is unrealistically small, and the p-value is artificially perfect, due to small cluster size 2 counties, don't use this, overfit the model 

did_fe_robust <- feols(DeathRate ~ Treat * Post | County + Year, data = combined_clean,
                       cluster = ~County)
summary(did_fe_robust)
```



## Placebo Test (Using Pre-Treatment Years Only)

```{r placebo}
# Subset data to pre-treatment years only
pre_data <- filter(data, year <= 2020)

# Fake "placebo" policy year in 2019
pre_data <- pre_data %>%
  mutate(placebo_post = ifelse(year >= 2019, 1, 0),
         placebo_did = placebo_post * treatment_group)

placebo_results <- lapply(outcomes, function(outcome) {
  model <- feols(as.formula(paste(outcome, "~ placebo_post * treatment_group | city + year")), data = pre_data, cluster = ~city)
  tidy(model) %>%
    mutate(outcome = outcome)
})

placebo_df <- bind_rows(placebo_results)
placebo_df
```

## Interpretation

- DiD estimates with fixed effects help control for city-specific and time-specific confounders.
- **Parallel trends plots** visually check the pre-treatment alignment of trends.
- **Placebo tests** ensure no "effect" is detected in pre-policy years, increasing internal validity.

